risk-perception

### 2019-10-24 risk-percption.Rmd
knitable document

```{r}
agents <- tibble(id = 1:5, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta))
```

@@

And payoffs for different crop types in different years.
```{r}
wheat_dry <- 0
wheat_wet <- 1.6
barley_dry <- 0.93
barley_wet <- 1.18

wheat_dry <- .2
wheat_wet <- .4
barley_dry <- 0.6
barley_wet <- .3
```

@@

The maximin strategy here is to plant barley all the time, but this entails a huge opportunity cost over planting wheat or a mix of the two

```{r, echo = FALSE}
plt_dat2 <- tibble(wheat_prop = seq(0, 1, .1),
       dry_exp = wheat_prop * wheat_dry + (1 - wheat_prop) * barley_dry,
      wet_exp = wheat_prop * wheat_wet + (1 - wheat_prop) * barley_wet) %>%
  gather(type, utility, dry_exp:wet_exp)
```

So the intersection point of the next plot is the mixed strategy equilbrium(indifference point) in a risk-free environment

```{r, echo = FALSE}
nash_eq <- tibble(wheat_prop = abs(barley_dry - barley_wet) / (abs(barley_dry - barley_wet) + abs(wheat_dry - wheat_wet)),
                  utility = wheat_prop * (wheat_dry + wheat_wet))

plt_dat2 %>%
  ggplot(aes(wheat_prop, utility)) +
  geom_hline(yintercept = nash_eq$utility, linetype = 2)+
  geom_vline(xintercept = nash_eq$wheat_prop, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = nash_eq, color = 'black', size = 3) +
  geom_text(data = nash_eq, 
            label = paste0('(', nash_eq$wheat_prop, ', ', nash_eq$utility, ')'),
            nudge_x = 0.1, nudge_y = 0.05) +
  theme_classic()

@@

Now create a population of agents with different subjective beliefs about the drought risk.
```{r}

test2 <- c((runif(25) < .2),(runif(25) < .5), runif(25) < .1)
tibble(time = 1:75,
       drought = test2) %>%
  mutate(period = round(time, -1)) %>%
  add_count(period) %>%
  filter(period >= 10) %>%
  group_by(period) %>%
  summarise(drought = sum(drought)) %>%
  ggplot(aes(period, drought)) +
  geom_col()
```

```{r}
learn <- function(agents, event) {
  agents %>%
    mutate(alpha = event + alpha,
           beta = beta + 1 - event,
           mu = alpha / (alpha + beta),
           estimate = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}

accumulate(test2, ~learn(.x, .y), .init = test) %>% bind_rows(.id = 'time') %>% qplot(as.numeric(time), estimate, data = ., geom = 'point')
```

```{r}
learn(test, TRUE)
```

The basic risk assessment algorithm. Let's look at the risk of a drought year being 0.2. In a wet year you get 3 wheat and 2 barley, and in a dry year you get 0 wheat and 1 barley. What does that look like?

@@

```{r}
test_dat <- expand_grid(wheat_prop = (0:10)/10,
       risk = seq(0, 1, 0.1)) %>%
  mutate(util_wet = (1 - risk) * (wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop)),
         util_dry = risk * (wheat_dry * wheat_prop + barley_dry * (1 - wheat_prop)),
         util_wheat = wheat_prop * (wheat_wet * (1 - risk) + wheat_dry *  risk),
         util_barley = (1 - wheat_prop) * (barley_wet* (1 - risk) + barley_dry * risk),
         util_payoff = util_dry + util_wet)
```

```{r}
test_dat %>%
  select(-(util_wheat:util_barley)) %>%
  gather(type, util, c(util_dry, util_wet, util_payoff)) %>%
    ggplot(aes(wheat_prop, util, color = type)) + geom_line() + facet_wrap(~risk) + geom_hline(yintercept = .5, alpha = .3, linetype = 2) + geom_vline(xintercept = .14, alpha = 0.3,linetype = 2)
```
```{r}
test_dat %>%
  select(-(util_wheat:util_barley)) %>%
  mutate(util_min = pmin(util_wet, util_dry))%>%
    ggplot(aes(util_payoff, util_min)) + geom_line() + facet_wrap(~risk) + geom_hline(yintercept = .3, alpha = .3, linetype = 2)
```

```{r}
test_dat %>%
  group_by(risk) %>%
  filter(payoff == max(payoff)) %>%
  ggplot(aes(risk, wheat_prop)) + geom_point()

test_dat %>%
  select(-(util_dry:util_wet)) %>%
  gather(type, util, util_wheat:util_barley) %>%
    ggplot(aes(wheat_prop, util, color = type)) + geom_line() + facet_wrap(~risk) + geom_vline(xintercept = .14)

test_dat %>%
  ggplot(aes(wheat_prop, util_wet + util_dry)) + geom_point() + facet_wrap(~risk)

calc_prop <- function(risk) {
  pb <- barley_wet * (1 - risk) - barley_dry * risk
  pw <- wheat_wet * (1 - risk) - wheat_dry * risk

  pb / (pb + pw) 
}

plot(calc_prop((0:10)/10))
```

```{r}
tibble(x)
```

```{r}
calcNE <- function(risk){
  (118 - 25 * risk) / (278 - 185 * risk)
}

(barley_wet - .2 * (barley_dry + wheat_dry + wheat_wet + barley_wet)) / (barley_wet + wheat_wet)

calcNE(1)

plot(1.6 * (1:10)/10 + 1.18 * (10:1)/10)

util_wet <- wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop)
util_dry <- wheat_dry * wheat_prop + barley_dry * (1 - wheat_prop)

tibble()

plot(seq(.1,1,.1) * (1 - seq(.1,1,.1)))

>>>>

```{r}
agents <- tibble(id = 1:100, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta))
```

@@

And payoffs for different crop types in different years.
```{r}
wheat_dry <- 0
wheat_wet <- 1.6
barley_dry <- 0.93
barley_wet <- 1.18
```

@@

The maximin strategy here is to plant barley all the time, but this entails a huge opportunity cost over planting wheat or a mix of the two. So the intersection point of the next plot is the mixed strategy equilbrium(indifference point) in a risk-free environment

```{r, echo = FALSE}
nash_eq <- tibble(wheat_prop = round((barley_wet - barley_dry) / (wheat_dry + barley_wet - (wheat_wet + barley_dry)), 2),
                  utility = round(wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop), 2))

tibble(wheat_prop = seq(0, 1, .1),
       dry_exp = wheat_prop * wheat_dry + (1 - wheat_prop) * barley_dry,
      wet_exp = wheat_prop * wheat_wet + (1 - wheat_prop) * barley_wet) %>%
  gather(type, utility, dry_exp:wet_exp) %>%
  ggplot(aes(wheat_prop, utility)) +
  geom_hline(yintercept = nash_eq$utility, linetype = 2)+
  geom_vline(xintercept = nash_eq$wheat_prop, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = nash_eq, color = 'black', size = 3) +
  geom_text(data = nash_eq, 
            label = paste0('(', nash_eq$wheat_prop, ', ', nash_eq$utility, ')'),
            nudge_x = 0.1, nudge_y = 0.05) +
  theme_classic()
```

```{r}
# farmer's ne
pmax(0, round((barley_wet - barley_dry) / (wheat_dry + barley_wet - (wheat_wet + barley_dry)), 2))

#nature's ne
round((barley_wet - wheat_wet) / (wheat_dry + barley_wet - (wheat_wet + barley_dry)), 2)
```

0.3111 is also the indifference point for the farmer with respect to crops. That is, if we imagine nature playing strategically, a drought would occur with probability 0.31111 because that would make nature indifferent to whether the farmer planted wheat or barley.

@@

Simulate a time series with varying drought frequency.
```{r}
droughts <- c((runif(25) < .1),(runif(25) < .5), runif(50) < .1)
```

Allow the agents to learn the drought risk over time.
```{r}
learn <- function(agents, event) {
  agents %>%
    mutate(alpha = event + alpha,
           beta = beta + 1 - event,
           mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

Run an experiment.
```{r}
simulation <- accumulate(droughts, ~learn(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

```{r}
ggplot(simulation, aes(time, risk)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

Think about the importance of the population's age structure, becuase that's going to impact the collective memory of drought.

```{r}
knitr::knit_exit()
```

@@

```{r}
test_dat <- expand_grid(wheat_prop = (0:10)/10,
       risk = seq(.1, .9, 0.1)) %>%
  mutate(util_wet = (1 - risk) * (wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop)),
         util_dry = risk * (wheat_dry * wheat_prop + barley_dry * (1 - wheat_prop)),
         util_wheat = wheat_prop * (wheat_wet * (1 - risk) + wheat_dry *  risk),
         util_barley = (1 - wheat_prop) * (barley_wet* (1 - risk) + barley_dry * risk),
         payoff_exp = util_dry + util_wet,
         payoff_max = pmax(util_wheat, util_barley))
```

```{r}
nash_eq2 <- tibble(risk = seq(0.1, 0.9, 0.1),
                   wheat_prop = pmax(round(((1 - risk) * barley_wet - risk * barley_dry) / (risk * wheat_dry + (1 - risk) * barley_wet - ((1 - risk) * wheat_wet + risk * barley_dry)), 2), 0),
                  utility = round(risk * wheat_dry * wheat_prop + risk * barley_dry * (1 - wheat_prop), 2))

test_dat %>%
  select(-(util_wheat:util_barley)) %>%
  gather(type, utility, c(util_dry, util_wet)) %>%
    ggplot(aes(wheat_prop, utility)) + 
  geom_line(aes(color = type)) + 
  facet_wrap(~risk) + 
  geom_hline(data = nash_eq2, aes(yintercept = utility), alpha = .3, linetype = 2) + 
  geom_vline(data = nash_eq2, aes(xintercept = wheat_prop), alpha = 0.3,linetype = 2) +
  geom_point(data = nash_eq2) +
  theme_classic()
```

```{r}
test_dat %>%
  select(-(util_dry:util_wet)) %>%
  gather(type, util, util_wheat:util_barley) %>%
  ggplot(aes(wheat_prop, util, color = type)) + 
  geom_line() + 
  facet_wrap(~risk) + 
  geom_vline(xintercept = nash_eq$wheat_prop)
```

Here we see that the mixed strategy that maximizes expected utility is always a pure strategy.
```{r}
test_dat %>%
  gather(type, utility, payoff_exp:payoff_max) %>%
  ggplot(aes(wheat_prop, utility)) + 
  geom_line(aes(color = type)) +
  facet_wrap(~risk)
```

So both the Wald and Savage regret criteria tell us to choose barley all the time.

If we're using the empirical frequency distirubtion of nature's moves, then we are playing a fictitious game. The strategy of planting the crop with the highest expected yield, given the frequency distribution of nature is known, will only work when nature is stationary.

We could think of other algorithms that would minimize the cumulative or average regret of a player over many years in a situation with nonstationary rainfall, but we have to remain within cognitive limits.

```{r}
# this is for nature
calcNE <- function(risk){
  (118 - 25 * risk) / (278 - 185 * risk)
}

plot(0:10, calcNE((0:10)/10))
```
